Comparing human and model-based forecasts of COVID-19 in Germany and Poland

### 2021-10-14 manuscript/manuscript.tex
update citations

Previous work has examined human predictions in various contexts, such as geopolitics (Tetlock et al. 2014; Atanasov et al. 2016), meta-science (Hoogeveen, Sarafoglou, and Wagenmakers 2020; {``Replication {Markets} -- {Reliable} Research Replicates\ldots you Can Bet on It.''} n.d.), sports (Servan-Schreiber et al. 2004) and epidemiology Recchia, Freeman, and Spiegelhalter (2021). Several prediction platforms (Metaculus 2020; {``Hypermind \textbar{} {Supercollective} Intelligence for Decision Makers''} n.d.; {``{CSET Foretell}''} n.d.) and prediction markets ({``{PredictIt}''} n.d.) have been created to collate expert and non-expert predictions. However, with the notable exception of Farrow et al., these crowd forecasts were not designed to be evaluated alongside model-based forecasts and usually follow their own (often binary) formats. Farrow et al.~found that humans could outperform computer models at predicting the 2014/15 and 2015/16 flu season in the US, a setting where the disease was well known and information about previous seasons was available. They tended to do slightly worse at predicting the 2014/15 outbreak of chikungunya in the Americas, a disease previously largely unobserved and unknown there and then.

@@

We created a crowd-sourced forecast (``crowd forecast''), elicited from participants through a web application (N. I. Bosse et al. 2020), as well as two semi-mechanistic models (``renewal model'' and ``convolution model'') informed by basic assumptions about COVID-19 epidemiology. Forecasts were created in real time over a period of 21 weeks from October 12th 2020 until March 1st 2021 and submitted to the German and Polish Forecast hub (Bracher, Wolffram, et al. 2021). All code and tools necessary to generate the forecasts and make a forecast submission are available in the \texttt{covid.german.forecasts} R package (N. Bosse et al. 2020). This repository also contains a record of all forecasts submitted to the German and Polish Forecast Hub. Forecasts were evaluated using a variety of scoring metrics and compared among each other and against an ensemble of all other models submitted to the Forecast Hub.

@@

Data on daily reported test positive cases and deaths linked to COVID-19 were provided by the organisers of the German and Polish forecast hub. Until December 14th 2020 these data were sourced from the European Centre for Disease Control (ECDC) ({``Download Historical Data (to 14 {December} 2020) on the Daily Number of New Reported {COVID}-19 Cases and Deaths Worldwide''} 2020). After ECDC stopped publishing daily data, observations were sourced from the Robert Koch Institute (RKI) and the Polish Ministry of Health for the remainder of the submission period ({``{RKI} - {Coronavirus SARS}-{CoV}-2 - {Aktueller Lage}-/{Situationsbericht} Des {RKI} Zu {COVID}-19''} n.d.). These data are subject to reporting artefacts (such as a delayed case reporting in Poland on the 24th November ({``Rozbieżności w statystykach koronawirusa. 22 tys. przypadków będą doliczone do ogólnej liczby wyników''} 16:07:56+0100)), changes in reporting over time and variation in testing regimes (e.g.~in Germany from the 11th of November on (Ärzteblatt 2020)).

@@

Our crowd forecasts were created as an ensemble of forecasts made by individual participants every week through a web application (\url{https://cmmid-lshtm.shinyapps.io/crowd-forecast/}). Forecasts were allowed until 12pm on Tuesdays, but participants were asked to only use information available until Monday. The application was built using the shiny and golem R packages (Fay et al. 2021; Chang et al. 2021) and is available in the \texttt{crowdforecastr} R package (N. I. Bosse et al. 2020). To make a forecast in the application participants could select a predictive distribution, with the default being log-normal, and adjust the median and the width of the uncertainty by either interacting with a figure showing their forecast or providing numerical values. The baseline shown was a repetition of the last known observation with constant uncertainty around it computed as the standard deviation of the last four changes in log observed forecasts (i.e.~as \texttt{sd(c(log(value4)\ -\ log(value3),\ log(value3)\ -\ log(value2),\ ...))}).
Our interface also allowed participants to view the observed data, and their forecasts, on a logarithmic scale and presented additional contextual COVID-19 data sourced from ({``{COVID}-19 {Data Explorer}''} n.d.). These data included e.g.~notifications of both test positive COVID-19 cases and COVID-19 linked deaths and the number of COVID-19 tests conducted over time.

@@

We used two models from the \texttt{EpiNow2} R package (version 1.3.3) as our model-based forecasts (Abbott, Hellewell, et al. 2020). The first of these models, here called ``renewal model,'' used the renewal equation (Fraser 2007) to predict reported cases and deaths. It estimated the effective reproduction number \(R_t\) (the average number of people each infected person is expected to infect in turn) and models future infections as a weighted sum of past infection multiplied by \(R_t\). \(R_t\) was assumed to stay constant beyond the forecast horizon, roughly corresponding to continuing the latest exponential trend in infections. On the 9th of November we altered the date when \(R_t\) was assumed to be constant from two weeks prior to the date of the forecast to the forecast date. Reported notifications were obtained by convolving predicted infections over data-based delay distributions (Abbott, Hellewell, et al. 2020; DOI n.d.; {``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid}-19 Outbreak in {England} \textbar{} {medRxiv}''} n.d.) to model the time between infection and report date or death. The renewal model was used to predict cases as well as deaths with forecasts being generated for each target separately, meaning that \(R_t\) was forecast once from reported cases and once from reported deaths. Death forecasts from the renewal model were therefore not informed by past cases. One submission of the renewal model on December 28th 2020 was delayed and therefore not included in the official Forecast hub ensemble.

The second model (``convolution model'') was only used to forecast deaths and was added later, starting December 7th 2020 (with the first forecast from December 7th suffering from a software bug and therefore disregarded in all further analyses). The convolution model was submitted, but never included in the official Forecast hub ensemble due to concerns that it could be too similar to the renewal model. The convolution model predicted deaths as a fraction of infected people who would die with some delay, by using a convolution of reported cases with a distribution that described the delay from case report to death and a scaling factor (the case-fatality ratio). Both the renewal and the convolution model used daily observations and assumed a negative binomial observation model with a multiplicative day-of-the-week effect (Abbott, Hellewell, et al. 2020). Line list data used to inform the prior for the delay from symptom onset to test positive case report or death in the model-based forecasts was sourced from (Xu et al., n.d.) with data available up to the 1st of August. All model fitting was done using Markov-chain Monte Carlo (MCMC) in stan (Stan Development Team 2020) with each location and forecast target being fit separately. More details are available in the supplementary information.

@@

@@

The prior for the generation time was sourced from (Ganyani et al. 2020) but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior (Lauer et al. 2020) for \(\xi_{O}\). This resulted in a gamma-distributed generation time with mean 3.6 days (standard deviation (SD) 0.7), and SD of 3.1 days (SD 0.8) for all estimates. We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data (Xu et al., n.d.; Abbott, Sherratt, et al. 2020) and combining the posteriors for the mean and standard deviation of the log-normal distribution (Abbott, Hellewell, et al. 2020; DOI n.d.; {``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid}-19 Outbreak in {England} \textbar{} {medRxiv}''} n.d.; Stan Development Team 2020).

@@

\leavevmode\vadjust pre{\hypertarget{ref-crowdforecastr}{}}%
Bosse, Nikos I., Sam Abbott, EpiForecasts, and Sebastian Funk. 2020. \emph{Crowdforecastr: Eliciting Crowd Forecasts in r Shiny}. \url{https://doi.org/10.5281/zenodo.4618519}.

\leavevmode\vadjust pre{\hypertarget{ref-covidgermanforecasts}{}}%
Bosse, Nikos, Sam Abbott, EpiForecasts, and Sebastian Funk. 2020. \emph{Covid.german.forecasts: Forecasting Covid-19 Related Metrics for the German/Poland Forecast Hub}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-CSETForetell}{}}%
{``{CSET Foretell}.''} n.d. Accessed October 13, 2021. \url{https://www.cset-foretell.com/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-doiCovid19TemporalVariation}{}}%
DOI, Authors Affiliations Published Not published yet. n.d. {``Covid-19: Temporal Variation in Transmission During the {COVID}-19 Outbreak.''} {Covid-19}. Accessed May 30, 2021. \url{https://epiforecasts.io/covid/}.

\leavevmode\vadjust pre{\hypertarget{ref-DownloadHistoricalData2020a}{}}%
{``Download Historical Data (to 14 {December} 2020) on the Daily Number of New Reported {COVID}-19 Cases and Deaths Worldwide.''} 2020. {European Centre for Disease Prevention and Control}. December 14, 2020. \url{https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-HypermindSupercollectiveIntelligence}{}}%
{``Hypermind \textbar{} {Supercollective} Intelligence for Decision Makers.''} n.d. {Hypermind}. Accessed October 13, 2021. \url{https://www.hypermind.com/en/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-PredictIt}{}}%
{``{PredictIt}.''} n.d. Accessed October 13, 2021. \url{https://www.predictit.org/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ReplicationMarketsReliable}{}}%
{``Replication {Markets} -- {Reliable} Research Replicates\ldots you Can Bet on It.''} n.d. Accessed October 13, 2021. \url{https://www.replicationmarkets.com/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-RKICoronavirusSARSCoV2a}{}}%
{``{RKI} - {Coronavirus SARS}-{CoV}-2 - {Aktueller Lage}-/{Situationsbericht} Des {RKI} Zu {COVID}-19.''} n.d. Accessed May 30, 2021. \url{https://www.rki.de/DE/Content/InfAZ/N/Neuartiges_Coronavirus/Situationsberichte/Gesamt.html}.

\leavevmode\vadjust pre{\hypertarget{ref-RozbieznosciStatystykachKoronawirusa0100}{}}%
{``Rozbieżności w statystykach koronawirusa. 22 tys. przypadków będą doliczone do ogólnej liczby wyników.''} 16:07:56+0100. 16:07:56+0100. \url{https://forsal.pl/lifestyle/zdrowie/artykuly/8017628,rozbieznosci-w-statystykach-koronawirusa-22-tys-przypadkow-beda-doliczone-do-ogolnej-liczby-wynikow.html}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-kraemer2020epidemiological}{}}%
Xu, Bo, Bernardo Gutierrez, Sarah Hill, Samuel Scarpino, Alyssa Loskill, Jessie Wu, Kara Sewalk, et al. n.d. {``Epidemiological Data from the nCoV-2019 Outbreak: Early Descriptions from Publicly Available Data.''} \url{http://virological.org/t/epidemiological-data-from-the-ncov-2019-outbreak-early-descriptions-from-publicly-available-data/337}.

>>>>

Previous work has examined human predictions in various contexts, such as geopolitics (Tetlock et al. 2014; Atanasov et al. 2016), meta-science (Hoogeveen, Sarafoglou, and Wagenmakers 2020; ReplicationMarkets 2020), sports (Servan-Schreiber et al. 2004) and epidemiology (Farrow et al. 2017; McAndrew and Reich 2020; Recchia, Freeman, and Spiegelhalter 2021). Several prediction platforms (Metaculus 2020; Hypermind 2021; CSET Foretell 2021) and prediction markets (PredictIt 2021) have been created to collate expert and non-expert predictions. However, with the notable exception of Farrow et al., these crowd forecasts were not designed to be evaluated alongside model-based forecasts and usually follow their own (often binary) formats. Farrow et al.~found that humans could outperform computer models at predicting the 2014/15 and 2015/16 flu season in the US, a setting where the disease was well known and information about previous seasons was available. They tended to do slightly worse at predicting the 2014/15 outbreak of chikungunya in the Americas, a disease previously largely unobserved and unknown there and then.

@@

We created a crowd-sourced forecast (``crowd forecast''), elicited from participants through a web application (Bosse et al. 2020b), as well as two semi-mechanistic models (``renewal model'' and ``convolution model'') informed by basic assumptions about COVID-19 epidemiology. Forecasts were created in real time over a period of 21 weeks from October 12th 2020 until March 1st 2021 and submitted to the German and Polish Forecast hub (Bracher, Wolffram, et al. 2021). All code and tools necessary to generate the forecasts and make a forecast submission are available in the \texttt{covid.german.forecasts} R package (Bosse et al. 2020a). This repository also contains a record of all forecasts submitted to the German and Polish Forecast Hub. Forecasts were evaluated using a variety of scoring metrics and compared among each other and against an ensemble of all other models submitted to the Forecast Hub.

@@

Data on daily reported test positive cases and deaths linked to COVID-19 were provided by the organisers of the German and Polish forecast hub. Until December 14th 2020 these data were sourced from the European Centre for Disease Control (ECDC) (ECDC 2020). After ECDC stopped publishing daily data, observations were sourced from the Robert Koch Institute (RKI) and the Polish Ministry of Health for the remainder of the submission period (RKI 2021). These data are subject to reporting artefacts (such as a delayed case reporting in Poland on the 24th November (Forsal.pl 2020)), changes in reporting over time and variation in testing regimes (e.g.~in Germany from the 11th of November on (Ärzteblatt 2020)).

@@

Our crowd forecasts were created as an ensemble of forecasts made by individual participants every week through a web application (\url{https://cmmid-lshtm.shinyapps.io/crowd-forecast/}). Forecasts were allowed until 12pm on Tuesdays, but participants were asked to only use information available until Monday. The application was built using the shiny and golem R packages (Fay et al. 2021; Chang et al. 2021) and is available in the \texttt{crowdforecastr} R package (Bosse et al. 2020b). To make a forecast in the application participants could select a predictive distribution, with the default being log-normal, and adjust the median and the width of the uncertainty by either interacting with a figure showing their forecast or providing numerical values. The baseline shown was a repetition of the last known observation with constant uncertainty around it computed as the standard deviation of the last four changes in log observed forecasts (i.e.~as \(\sigma(log(value4) - log(value3), log(value3) - log(value2), \ldots )\)).
Our interface also allowed participants to view the observed data, and their forecasts, on a logarithmic scale and presented additional contextual COVID-19 data sourced from ({``{COVID}-19 {Data Explorer}''} n.d.). These data included e.g.~notifications of both test positive COVID-19 cases and COVID-19 linked deaths and the number of COVID-19 tests conducted over time.

@@

We used two models from the \texttt{EpiNow2} R package (version 1.3.3) as our model-based forecasts (Abbott, Hellewell, et al. 2020). The first of these models, here called ``renewal model,'' used the renewal equation (Fraser 2007) to predict reported cases and deaths. It estimated the effective reproduction number \(R_t\) (the average number of people each infected person is expected to infect in turn) and models future infections as a weighted sum of past infection multiplied by \(R_t\). \(R_t\) was assumed to stay constant beyond the forecast horizon, roughly corresponding to continuing the latest exponential trend in infections. On the 9th of November we altered the date when \(R_t\) was assumed to be constant from two weeks prior to the date of the forecast to the forecast date. Reported notifications were obtained by convolving predicted infections over data-based delay distributions (Abbott, Hellewell, et al. 2020; epiforecasts.io/covid n.d.; Sherratt et al. 2021) to model the time between infection and report date or death. The renewal model was used to predict cases as well as deaths with forecasts being generated for each target separately, meaning that \(R_t\) was forecast once from reported cases and once from reported deaths. Death forecasts from the renewal model were therefore not informed by past cases. One submission of the renewal model on December 28th 2020 was delayed and therefore not included in the official Forecast hub ensemble.

The second model (``convolution model'') was only used to forecast deaths and was added later, starting December 7th 2020 (with the first forecast from December 7th suffering from a software bug and therefore disregarded in all further analyses). The convolution model was submitted, but never included in the official Forecast hub ensemble due to concerns that it could be too similar to the renewal model. The convolution model predicted deaths as a fraction of infected people who would die with some delay, by using a convolution of reported cases with a distribution that described the delay from case report to death and a scaling factor (the case-fatality ratio). Both the renewal and the convolution model used daily observations and assumed a negative binomial observation model with a multiplicative day-of-the-week effect (Abbott, Hellewell, et al. 2020). Line list data used to inform the prior for the delay from symptom onset to test positive case report or death in the model-based forecasts was sourced from (Xu et al. 2020) with data available up to the 1st of August. All model fitting was done using Markov-chain Monte Carlo (MCMC) in stan (Stan Development Team 2020) with each location and forecast target being fit separately. More details are available in the supplementary information.

@@

\hypertarget{the-crowdforecasting-app}{%
\subsection{The crowdforecasting app}\label{the-crowdforecasting-app}}

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../crowd-forecast/Screenshot-forecasting-app} \caption{Screenshot of the crowdforecasting app used to elicit predictions (made in June 2021). }\label{fig:screenshot}
\end{figure}

@@

The prior for the generation time was sourced from (Ganyani et al. 2020) but refit using a log-normal incubation period with a mean of 5.2 days (SD 1.1) and SD of 1.52 days (SD 1.1) with this incubation period also being used as a prior (Lauer et al. 2020) for \(\xi_{O}\). This resulted in a gamma-distributed generation time with mean 3.6 days (standard deviation (SD) 0.7), and SD of 3.1 days (SD 0.8) for all estimates. We estimated the delay between symptom onset and case report or death required to convolve latent infections to observations by fitting an integer adjusted log-normal distribution to 10 subsampled bootstraps of a public linelist for cases in Germany from April 2020 to June 2020 with each bootstrap using 1\% or 1769 samples of the available data (Xu et al. 2020; Abbott, Sherratt, et al. 2020) and combining the posteriors for the mean and standard deviation of the log-normal distribution (Abbott, Hellewell, et al. 2020; epiforecasts.io/covid n.d.; {``Evaluating the Use of the Reproduction Number as an Epidemiological Tool, Using Spatio-Temporal Trends of the {Covid}-19 Outbreak in {England} \textbar{} {medRxiv}''} n.d.; Stan Development Team 2020).

@@

\leavevmode\vadjust pre{\hypertarget{ref-covidgermanforecasts}{}}%
Bosse, Nikos I., Sam Abbott, EpiForecasts, and Sebastian Funk. 2020a. \emph{Covid.german.forecasts: Forecasting Covid-19 Related Metrics for the German/Poland Forecast Hub}.

\leavevmode\vadjust pre{\hypertarget{ref-crowdforecastr}{}}%
---------. 2020b. \emph{Crowdforecastr: Eliciting Crowd Forecasts in r Shiny}. \url{https://doi.org/10.5281/zenodo.4618519}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-csetforetellCSETForetell2021}{}}%
CSET Foretell. 2021. {``{CSET Foretell}.''} 2021. \url{https://www.cset-foretell.com/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ecdcDownloadHistoricalData2020}{}}%
ECDC. 2020. {``Download Historical Data (to 14 {December} 2020) on the Daily Number of New Reported {COVID}-19 Cases and Deaths Worldwide.''} {European Centre for Disease Prevention and Control}. December 14, 2020. \url{https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide}.

\leavevmode\vadjust pre{\hypertarget{ref-epiforecasts.ioux2fcovidCovid19TemporalVariation}{}}%
epiforecasts.io/covid. n.d. {``Covid-19: Temporal Variation in Transmission During the {COVID}-19 Outbreak.''} {Covid-19}. Accessed May 30, 2021. \url{https://epiforecasts.io/covid/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-forsal.plRozbieznosciStatystykachKoronawirusa2020}{}}%
Forsal.pl. 2020. {``Rozbieżności w statystykach koronawirusa. 22 tys. przypadków będą doliczone do ogólnej liczby wyników.''} 2020. \url{https://forsal.pl/lifestyle/zdrowie/artykuly/8017628,rozbieznosci-w-statystykach-koronawirusa-22-tys-przypadkow-beda-doliczone-do-ogolnej-liczby-wynikow.html}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hypermindHypermindSupercollectiveIntelligence2021}{}}%
Hypermind. 2021. {``Hypermind \textbar{} {Supercollective} Intelligence for Decision Makers.''} {Hypermind}. 2021. \url{https://www.hypermind.com/en/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-predictitPredictIt2021}{}}%
PredictIt. 2021. {``{PredictIt}.''} 2021. \url{https://www.predictit.org/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-replicationmarketsReplicationMarketsReliable2020}{}}%
ReplicationMarkets. 2020. {``Replication {Markets} -- {Reliable} Research Replicates\ldots you Can Bet on It.''} 2020. \url{https://www.replicationmarkets.com/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rkiRKICoronavirusSARSCoV22021}{}}%
RKI. 2021. {``{RKI} - {Coronavirus SARS}-{CoV}-2 - {Aktueller Lage}-/{Situationsbericht} Des {RKI} Zu {COVID}-19.''} 2021. \url{https://www.rki.de/DE/Content/InfAZ/N/Neuartiges_Coronavirus/Situationsberichte/Gesamt.html}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-sherrattExploringSurveillanceData2021}{}}%
Sherratt, Katharine, Sam Abbott, Sophie R. Meakin, Joel Hellewell, James D. Munday, Nikos Bosse, CMMID Covid-19 working Group, Mark Jit, and Sebastian Funk. 2021. {``Exploring Surveillance Data Biases When Estimating the Reproduction Number: With Insights into Subpopulation Transmission of {Covid}-19 in {England},''} March, 2020.10.18.20214585. \url{https://doi.org/10.1101/2020.10.18.20214585}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kraemer2020epidemiological}{}}%
Xu, Bo, Bernardo Gutierrez, Sarah Hill, Samuel Scarpino, Alyssa Loskill, Jessie Wu, Kara Sewalk, et al. 2020. {``Epidemiological Data from the nCoV-2019 Outbreak: Early Descriptions from Publicly Available Data.''} \url{http://virological.org/t/epidemiological-data-from-the-ncov-2019-outbreak-early-descriptions-from-publicly-available-data/337}.
